Counterfactual Training: Teaching Models Plausible and Actionable Explanations

### 2025-03-10 paper/preprint/paper.tex

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\newcounter{quartocalloutnteno}
\newcommand{\quartocalloutnte}[1]{\refstepcounter{quartocalloutnteno}\label{#1}}

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern machine learning (ML) models are tasked with learning these
representations from scratch, guided by narrow objectives such as
predictive accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow,
Bengio, and Courville 2016}). Modern advances in computing have made it
possible to provide such models with ever greater degrees of freedom to
achieve that task, which has often led them to outperform traditionally
more parsimonious models. Unfortunately, in doing so they also learn
increasingly complex and highly sensitive representations that we can no
longer easily interpret.

This trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, Szegedy et al. (\citeproc{ref-szegedy2013intriguing}{2013})
showed that artificial neural networks (ANN) are sensitive to
adversarial examples: counterfactuals of model inputs that yield vastly
different model predictions despite being ``imperceptible'' in that they
are semantically indifferent from their factual counterparts. Despite
partially effective mitigation strategies such as \textbf{adversarial
training} (\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow,
Shlens, and Szegedy 2014}), truly robust deep learning (DL) remains
unattainable even for models that are considered shallow by today's
standards (\citeproc{ref-kolter2023keynote}{Kolter 2023}).

Part of the problem is that high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020})\footnote{For
  clarity: we follow standard ML convention in using ``degrees of
  freedom'' to refer to the number of parameters estimated from data.}.
Based purely on predictive performance, these solutions may seem to
provide compelling explanations for the data, when in fact they are
based on purely associative, semantically meaningless patterns. This
poses two related challenges: firstly, it makes these models inherently
opaque, since humans cannot simply interpret what type of explanation
the complex learned representations correspond to; secondly, even if we
could resolve the first challenge, it is not obvious how to mitigate
models from learning representations that correspond to meaningless and
implausible explanations.

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has already occurred
in practice and caused real harm to people who were affected adversely
and often unfairly by automated decision-making systems (ADMS) involving
opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016}). Effective
XAI tools can aide us in monitoring models and providing recourse to
individuals to turn adverse outcomes (e.g.~``loan application
rejected'') into positive ones (``application accepted''). Wachter,
Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}) propose
\textbf{counterfactual explanations} as an effective approach to achieve
this: they explain how factual inputs need to change in order for some
fitted model to produce some desired output, typically involving minimal
perturbations.

To our surprise, the second challenge has not yet attracted any
consolidated research effort. Specifically, there has been no concerted
effort towards improving model \textbf{explainability}, which we define
here as the degree to which learned representations correspond to
explanations that are interpretable and deemed \textbf{plausible} by
humans (see Definition~\ref{def-explainability}). Instead, the choice
has typically been to improve the capacity of XAI tools to identify the
subset explanations that are both plausible and valid for any given
model, independent of whether the learned representations are also
compatible with implausible explanations
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
Fortunately, recent findings indicate that explainability can arise as
byproduct of regularization techniques aimed at other objectives such as
robustness, generalization and generative capacity Altmeyer et al.
(\citeproc{ref-altmeyer2024faithful}{2024}).

Building on these findings, we introduce \textbf{counterfactual
training}: a novel regularization technique geared explicitly towards
aligning model representations with plausible explanations. Our
contributions are as follows:

@@

Despite limitations of our approach discussed in
Section~\ref{sec-discussion}, we conclude that counterfactual training
provides a practical framework for researchers and practitioners
interested in making opaque models more trustworthy
Section~\ref{sec-conclusion}. We also believe that this work serves as
an opportunity for XAI researchers to reevaluate the premise of
improving XAI tools without improving models.

@@

To the best of our knowledge, our proposed framework for counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow Augustin, Meinke, and Hein
(\citeproc{ref-augustin2020adversarial}{2020}) in tieing the concept of
explainability to the quality of counterfactual explanations that we can
generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e.~adversarial examples),
AT effectively removes some implausible explanations from the solution
space.

@@

This interpretation of the link between explainability through
counterfactuals on one side, and robustness to adversarial examples on
the other, is backed by empirical evidence. Sauer and Geiger
(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrate that using
counterfactual images during classifier training improves model
robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate the
augmenting the training data of image classifiers can improve
generalization. Teney, Abbasnedjad, and Hengel
(\citeproc{ref-teney2020learning}{2020}) propose an approach using
counterfactuals in training that does not rely on data augmentation:
they argue that counterfactual pairs typically already exist in training
datasets. Specifically, their approach relies on, firstly, identifying
similar input samples with different annotations and, secondly, ensuring
that the gradient of the classifier aligns with the vector between pairs
of counterfactual inputs using the cosine distance as a loss function.

@@

There have also been several attempts at formalizing the relationship
between counterfactual explanations (CE) and adversarial examples (AE).
Pointing to clear similarities in how CE and AE are generated,
Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) makes the
case for jointly studying the opaqueness and robustness problem in
representation learning. Formally, AE can be seen as the subset of CE,
for which misclassification is achieved
(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) show that
CE and AE are equivalent under certain conditions and derive theoretical
upper bounds on the distances between them.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of post-hoc counterfactual explanations. Firstly,
Ross, Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024})
propose a way to train models that are guaranteed to provide recourse
for individuals to move from an adverse outcome to some positive target
class with high probability. The approach proposed by Ross, Lakkaraju,
and Bastani (\citeproc{ref-ross2021learning}{2024}) builds on
adversarial training, where in this context susceptibility to targeted
adversarial examples for the positive class is explicitly induced. The
proposed method allows for imposing a set of actionability constraints
ex-ante: for example, users can specify that certain features
(e.g.~\emph{age}, \emph{gender}, \ldots) are immutable. Secondly, Guo,
Nguyen, and Yadav (\citeproc{ref-guo2023counternet}{2023}) are the first
to propose an end-to-end training pipeline that includes counterfactual
explanations as part of the training procedure. In particular, they
propose a specific network architecture that includes a predictor and CE
generator network, where the parameters of the CE generator network are
learnable. Counterfactuals are generated during each training iteration
and fed back to the predictor network. In contrast to Guo, Nguyen, and
Yadav (\citeproc{ref-guo2023counternet}{2023}), we impose no
restrictions on the neural network architecture at all.

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CE,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
is a necessary but insufficient condition for using CE to provide
algorithmic recourse (AR) to individuals affected by opaque models in
practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this complies with the definition of
plausibility if 40-year-old individuals are in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant since \emph{age} is not a
mutable feature. For our intents and purposes, counterfactual training
aims at improving model explainability by aligning models with
counterfactuals that meet both desiderata, plausibility and
actionability. Formally, we define explainability as follows:

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}). For our purposes, we
let \(T\) vary and consider the counterfactual search as converged as
soon as the predicted probability for the target class has reached a
pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\)\footnote{For
  detailed background information on gradient-based counterfactual
  search and convergence see Section~\ref{sec-app-ce}.}.

To train models with high explainability as defined in
Definition~\ref{def-explainability}, we propose to leverage
counterfactuals in the following objective,

@@

where \(\text{yloss}(\cdot)\) is any conventional classification loss
that induces discriminative performance (e.g.~cross-entropy). The two
additional components in Equation~\ref{eq-obj} are explained in more
detail below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: 1) the contrastive
divergence, \(\text{div}(\cdot)\), between counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, 2) the
adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between
the different components can be governed by adjusting the strengths of
the penalties \(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).

@@

Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) observe that
any classifier can be re-interpreted as a joint energy-based model (JEM)
that learns to discriminate output classes conditional on inputs and
generate inputs. They show that JEMs can be trained to perform well at
both tasks by directly maximizing the joint log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first factor can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimize the
contrastive divergence between samples drawn from
\(p_\theta(\mathbf{x})\) and training observations, i.e.~samples from
\(p(\mathbf{x})\).

@@

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any counterfactual explainer that relies on gradient
descent in the feature space for optimization
(\citeproc{ref-altmeyer2023explaining}{Altmeyer, Deursen, et al. 2023}).
In this context, domain constraints can be imposed by simply projecting
counterfactuals back to the specified domain, if the previous gradient
step resulted in updated feature values that were out-of-domain.
Mutability constraints can similarly be enforced by setting partial
derivatives to zero to ensure that features are only mutated in the
allowed direction, if at all.

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to align model representations with user requirements. Through
their effect on \(\mathbf{x}^\prime\), both types of constraints
influence model outcomes through Equation~\ref{eq-div}. Here it is
crucial that we avoid penalizing implausibility that arises due to
mutability constraints. For any mutability-constrained feature \(d\)
this can be achieved by enforcing
\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
violate mutability constraints. Specifically, we set
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if

@@

From a Bayesian perspective, setting
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
Intuitively, we think of this simply in terms ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This in turn results in lower overall sensitivity to immutable
features, which we demonstrate empirically for different classifiers in
Section~\ref{sec-experiments}. Under certain conditions, this results
holds theoretically{[}For the proof, see the supplementary appendix.{]}:

@@

Let
\(f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})\)
denote a linear classifier with softmax activation \(\mathcal{S}\)
(i.e.~\emph{multinomial logistic regression}) where
\(y\in\{1,...,K\}=\mathcal{K}\) and \(\mathbf{x} \in \mathbb{R}^D\). If
we assume multivariate Gaussian class densities with common diagonal
covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\),
then protecting an immutable feature from the contrastive divergence
penalty (Equation~\ref{eq-div}) will result in lower classifier
sensitivity to that feature relative to the remaining features, provided
that at least one of those is mutable and discriminative.

@@

Suppose we are interested in predicting the likelihood that loan
applicants default on their credit. We have access to historical data on
previous loan takers comprised of a binary outcome variable
(\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
1) the subjects' \emph{age}, which we define as immutable, and 2) the
subjects' existing level of \emph{debt}, which we define as mutable.

@@

\FloatBarrier

>>>>

\usepackage{arxiv}
\usepackage{orcidlink}
\usepackage{amsmath}
\usepackage[T1]{fontenc}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\usepackage{amsthm}
\theoremstyle{plain}
\newtheorem{corollary}{Research Question}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\theoremstyle{plain}
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{remark}
\AtBeginDocument{\renewcommand*{\proofname}{Proof}}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\newtheorem{refremark}{Remark}[section]
\newtheorem{refsolution}{Solution}[section]
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

Today's prominence of artificial intelligence (AI) has largely been
driven by advances in \textbf{representation learning}: instead of
relying on features and rules that are carefully hand-crafted by humans,
modern machine learning (ML) models are tasked with learning the
representations directly from data, guided by narrow objectives such as
predictive accuracy (\citeproc{ref-goodfellow2016deep}{I. Goodfellow,
Bengio, and Courville 2016}). Modern advances in computing have made it
possible to provide such models with ever growing degrees of freedom to
achieve that task, which has often led them to outperform traditionally
more parsimonious models. Unfortunately, in doing so, the models learn
increasingly complex and highly sensitive representations that we can no
longer easily interpret.

The trend towards complexity for the sake of performance has come under
serious scrutiny in recent years. At the very cusp of the deep learning
revolution, Szegedy et al. (\citeproc{ref-szegedy2013intriguing}{2013})
showed that artificial neural networks (ANN) are sensitive to
adversarial examples: counterfactuals of model inputs that yield vastly
different model predictions despite being ``imperceptible'' in that they
are semantically indifferent from their factual counterparts. Although
some partially effective mitigation strategies have been proposed, for
example \textbf{adversarial training}
(\citeproc{ref-goodfellow2014explaining}{I. J. Goodfellow, Shlens, and
Szegedy 2014}), truly robust deep learning (DL) remains unattainable
even for models that are considered shallow by today's standards
(\citeproc{ref-kolter2023keynote}{Kolter 2023}).

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives (\citeproc{ref-wilson2020case}{Wilson 2020}).\footnote{We
  follow the standard ML convention, where ``degrees of freedom'' refer
  to the number of parameters estimated from data.} Indeed, recent work
on the so called ``lottery tickets'' suggests that modern neural
networks can be pruned by up to 90\% while preserving their predictive
performance (\citeproc{ref-frankle2018lottery}{Frankle and Carbin 2019})
and generalizability (\citeproc{ref-morcos2019success}{Morcos et al.
2019}). Similarly, Zhang et al.
(\citeproc{ref-zhang2021understanding}{2021}) showed that
state-of-the-art neural networks are so expressive that they can fit
randomly labeled data. Thus, looking at the predictive performance, the
solutions may seem to provide compelling explanations for the data, when
in fact they are based on purely associative, semantically meaningless
patterns. This poses two related challenges. Firstly, there is no
dependable way to verify if such complex representations correspond to
meaningful and plausible explanations. Secondly, even if we could
resolve the first challenge, it remains undecided how to ensure that
models can \emph{only} learn valuable explanations.

The first challenge has attracted an abundance of research on
\textbf{explainable AI} (XAI) which aims to develop tools to derive
explanations from complex model representations. This can mitigate a
scenario in which we deploy opaque models and blindly rely on their
predictions. On countless occasions, this scenario has occurred in
practice and caused real harm to people who were affected adversely and
often unfairly by automated decision-making (ADM) systems involving
opaque models (\citeproc{ref-oneil2016weapons}{O'Neil 2016};
\citeproc{ref-mcgregor2021preventing}{McGregor 2021}). Effective XAI
tools can aid us in monitoring models and providing recourse to
individuals to turn adverse outcomes (e.g., ``loan application
rejected'') into positive ones (e.g., ``application accepted'').
Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}) propose
\textbf{counterfactual explanations} (CE) as an effective approach to
achieve this goal: CEs explain how factual inputs need to change in
order for some fitted model to produce some desired output, typically
involving minimal perturbations.

To our surprise, the second challenge has not yet attracted any major
consolidated research effort. Specifically, there has been no concerted
effort towards improving improving models' explanatory capacity, which
we will henceforth simply call ``explainability'', defined as the degree
to which learned representations correspond to explanations that are
interpretable and deemed \textbf{plausible} by humans (see
Definition~\ref{def-explainability}). Instead, the choice has typically
been to improve the ability of XAI tools to identify the subset
explanations that are both plausible and valid for any given model,
independent of whether the learned representations are also compatible
with implausible explanations
(\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).
Fortunately, recent findings indicate that explainability can arise as
byproduct of regularization techniques aimed at other objectives such as
robustness, generalization, and generative capacity
(\citeproc{ref-schut2021generating}{Schut et al. 2021};
\citeproc{ref-augustin2020adversarial}{Augustin, Meinke, and Hein 2020};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}).

Building on these findings, we introduce \textbf{counterfactual
training}: a novel training regime explicitly geared towards aligning
model representations with plausible explanations. Our contributions are
as follows:

@@

Despite some limitations of our approach discussed in
Section~\ref{sec-discussion}, we conclude in
Section~\ref{sec-conclusion} that counterfactual training provides a
practical framework for researchers and practitioners interested in
making opaque models more trustworthy. We also believe that this work
serves as an opportunity for XAI researchers to re-evaluate the trend of
improving XAI tools without improving the underlying models.

@@

To the best of our knowledge, our proposed framework of counterfactual
training represents the first attempt to use counterfactual explanations
during training to improve model explainability. In high-level terms, we
define model explainability as the extent to which valid explanations
derived for an opaque model are also deemed plausible with respect to
the underlying data and stakeholder requirements. To make this more
concrete, we follow Augustin, Meinke, and Hein
(\citeproc{ref-augustin2020adversarial}{2020}) in tying the concept of
explainability to the quality of counterfactual explanations that we can
generate for a given model. The authors show that counterfactual
explanations---understood here as minimal input perturbations that yield
some desired model prediction---are generally more meaningful if the
underlying model is more robust to adversarial examples. We can make
intuitive sense of this finding when looking at adversarial training
(AT) through the lens of representation learning with high degrees of
freedom: by inducing models to ``unlearn'' representations that are
susceptible to worst-case counterfactuals (i.e., adversarial examples),
AT effectively removes some implausible explanations from the solution
space.

@@

This interpretation of the link between explainability through
counterfactuals on one side, and robustness to adversarial examples on
the other, is backed by empirical evidence. Sauer and Geiger
(\citeproc{ref-sauer2021counterfactual}{2021}) demonstrate that using
counterfactual images during classifier training improves model
robustness. Similarly, Abbasnejad et al.
(\citeproc{ref-abbasnejad2020counterfactual}{2020}) argue that
counterfactuals represent potentially useful training data in machine
learning, especially in supervised settings where inputs may be
reasonably mapped to multiple outputs. They, too, demonstrate the
augmenting the training data of image classifiers can improve
generalization. Teney, Abbasnedjad, and Hengel
(\citeproc{ref-teney2020learning}{2020}) propose an approach using
counterfactuals in training that does not rely on data augmentation:
they argue that counterfactual pairs typically already exist in training
datasets. Specifically, their approach relies on identifying similar
input samples with different annotations and ensuring that the gradient
of the classifier aligns with the vector between such pairs of
counterfactual inputs using the cosine distance as the loss function.

@@

There have also been several attempts at formalizing the relationship
between counterfactual explanations and adversarial examples (AE).
Pointing to clear similarities in how CE and AE are generated,
Freiesleben (\citeproc{ref-freiesleben2022intriguing}{2022}) makes the
case for jointly studying the opaqueness and robustness problem in
representation learning. Formally, AE can be seen as the subset of CE
for which misclassification is achieved
(\citeproc{ref-freiesleben2022intriguing}{Freiesleben 2022}). Similarly,
Pawelczyk et al. (\citeproc{ref-pawelczyk2022exploring}{2022}) show that
CE and AE are equivalent under certain conditions and derive theoretical
upper bounds on the distances between them.

Two recent works are closely related to ours in that they use
counterfactuals during training with the explicit goal of affecting
certain properties of post-hoc counterfactual explanations. Firstly,
Ross, Lakkaraju, and Bastani (\citeproc{ref-ross2021learning}{2024})
propose a way to train models that are guaranteed to provide recourse
for individuals to move from an adverse outcome to some positive target
class with high probability. Their approach builds on adversarial
training, where in this context susceptibility to targeted adversarial
examples for the positive class is explicitly induced. The proposed
method allows for imposing a set of actionability constraints ex-ante:
for example, users can specify that certain features (e.g., \emph{age},
\emph{gender}, \ldots) are immutable. Secondly, Guo, Nguyen, and Yadav
(\citeproc{ref-guo2023counternet}{2023}) are the first to propose an
end-to-end training pipeline that includes counterfactual explanations
as part of the training procedure. In particular, they propose a
specific network architecture that includes a predictor and CE generator
network, where the parameters of the CE generator network are learnable.
Counterfactuals are generated during each training iteration and fed
back to the predictor network. In contrast to Guo, Nguyen, and Yadav
(\citeproc{ref-guo2023counternet}{2023}), we impose no restrictions on
the neural network architecture at all.

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit objective of aligning representations with plausible
explanations that comply with user requirements. In the context of CEs,
plausibility has broadly been defined as the degree to which
counterfactuals comply with the underlying data generating process
(\citeproc{ref-poyiadzi2020face}{Poyiadzi et al. 2020};
\citeproc{ref-guidotti2022counterfactual}{Guidotti 2022};
\citeproc{ref-altmeyer2024faithful}{Altmeyer et al. 2024}). Plausibility
is a necessary but insufficient condition for using CEs to provide
algorithmic recourse (AR) to individuals affected by opaque models in
practice. This is because for recourse recommendations to be
\textbf{actionable}, they need to not only result in plausible
counterfactuals but also be attainable. A plausible CE for a rejected
20-year-old loan applicant, for example, might reveal that their
application would have been accepted, if only they were 20 years older.
Ignoring all other features, this would comply with the definition of
plausibility if 40-year-old individuals were in fact more credit-worthy
on average than young adults. But of course this CE does not qualify for
providing actionable recourse to the applicant since \emph{age} is not a
(directly) mutable feature. For our intents and purposes, counterfactual
training aims to improve model explainability by aligning models with
counterfactuals that meet both desiderata, plausibility and
actionability. Formally, we define explainability as follows:

@@

Let \(\mathbf{x}_t^\prime\) for \(t=0,...,T\) denote a counterfactual
explanation generated through gradient descent over \(T\) iterations as
initially proposed by Wachter, Mittelstadt, and Russell
(\citeproc{ref-wachter2017counterfactual}{2017}). For our purposes, we
let \(T\) vary and consider the counterfactual search as converged as
soon as the predicted probability for the target class has reached a
pre-determined threshold, \(\tau\):
\(\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}^\prime))[y^+] \geq \tau\),
where \(\mathcal{S}\) is the softmax function.\footnote{For detailed
  background information on gradient-based counterfactual search and
  convergence see \textbf{?@sec-app-ce}.}

To train models with high explainability as defined in
Definition~\ref{def-explainability}, we propose to leverage
counterfactuals in the following objective:

@@

where \(\text{yloss}(\cdot)\) is any conventional classification loss
that induces discriminative performance (e.g., cross-entropy). The two
additional components in Equation~\ref{eq-obj} are explained in more
detail below. For now, they can be sufficiently described as inducing
explainability directly and indirectly by penalizing: (1) the
contrastive divergence, \(\text{div}(\cdot)\), between counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). The tradeoff between
the different components can be governed by adjusting the strengths of
the penalties \(\lambda_{\text{div}}\) and \(\lambda_{\text{adv}}\).

@@

Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) observe that
any classifier can be re-interpreted as a joint energy-based model (JEM)
that learns to discriminate output classes conditional on inputs and
generate inputs. They show that JEMs can be trained to perform well at
both tasks by directly maximizing the joint log-likelihood factorized as
\(\log p_\theta(\mathbf{x},\mathbf{y})=\log p_\theta(\mathbf{y}|\mathbf{x}) + \log p_\theta(\mathbf{x})\).
The first factor can be optimized using conventional cross-entropy as in
Equation~\ref{eq-obj}. To optimize \(\log p_\theta(\mathbf{x})\)
Grathwohl et al. (\citeproc{ref-grathwohl2020your}{2020}) minimize the
contrastive divergence between samples drawn from
\(p_\theta(\mathbf{x})\) and training observations, i.e., samples from
\(p(\mathbf{x})\).

@@

Many existing counterfactual explainers support domain and mutability
constraints out-of-the-box. In fact, both types of constraints can be
implemented for any counterfactual explainer that relies on gradient
descent in the feature space for optimization
(\citeproc{ref-altmeyer2023explaining}{Altmeyer, van Deursen, and Liem
2023}). In this context, domain constraints can be imposed by simply
projecting counterfactuals back to the specified domain, if the previous
gradient step resulted in updated feature values that were
out-of-domain. Mutability constraints can similarly be enforced by
setting partial derivatives to zero to ensure that features are only
mutated in the allowed direction, if at all.

Since actionability constraints are binding at test time, we should also
impose them when generating \(\mathbf{x}^\prime\) during each training
iteration to align model representations with user requirements. Through
their effect on \(\mathbf{x}^\prime\), both types of constraints
influence model outcomes through Equation~\ref{eq-div}. Here it is
crucial that we avoid penalizing implausibility that arises due to
mutability constraints. For any mutability-constrained feature \(d\)
this can be achieved by enforcing
\(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever perturbing
\(\mathbf{x}^\prime[d]\) in the direction of \(\mathbf{x}[d]\) would
violate mutability constraints. Specifically, we set
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:

@@

From a Bayesian perspective, setting
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
Intuitively, we think of this simply in terms ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This in turn results in lower overall sensitivity to immutable
features, which we demonstrate empirically for different classifiers in
Section~\ref{sec-experiments}. Under certain conditions, this results
holds theoretically:\footnote{For the proof, see the supplementary
  appendix.}

@@

Let
\(f_\theta(\mathbf{x})=\mathcal{S}(\mathbf{M}_\theta(\mathbf{x}))=\mathcal{S}(\Theta\mathbf{x})\)
denote a linear classifier with softmax activation \(\mathcal{S}\)
(i.e., \emph{multinomial logistic regression}) where
\(y\in\{1,...,K\}=\mathcal{K}\) and \(\mathbf{x} \in \mathbb{R}^D\). If
we assume multivariate Gaussian class densities with common diagonal
covariance matrix \(\Sigma_k=\Sigma\) for all \(k \in \mathcal{K}\),
then protecting an immutable feature from the contrastive divergence
penalty (Equation~\ref{eq-div}) will result in lower classifier
sensitivity to that feature relative to the remaining features, provided
that at least one of those is mutable and discriminative.

@@

Suppose we are interested in predicting the likelihood that loan
applicants default on their credit. We have access to historical data on
previous loan takers comprised of a binary outcome variable
(\(y\in\{1=\text{default},2=\text{no default}\}\)) two input features:
(1) the subjects' \emph{age}, which we define as immutable, and (2) the
subjects' existing level of \emph{debt}, which we define as mutable.

@@
